### 2014-06-27 manuscript/manuscript.tex
Final edits.

Noncoding RNAs are increasingly recognized as integral to a wide range
of biological processes, including core processes such as translation,
gene regulation, host-pathogen interactions and environmental
sensing. While bacterial and archaeal genomics is now a mature field,
our capacity to identify noncoding RNA elements in genomes is hampered
by the difficulty of de novo identification. The emergence of new
technologies for characterizing transcriptome outputs, notably
RNA-seq, are improving noncoding RNA identification and expression
quantification. However, a major challenge is to robustly distinguish
functional outputs from transcriptional noise. To establish whether
annotation of existing transcriptome data has effectively captured all
functional outputs, we analysed 400 publicly available RNA-seq
datasets spanning 40 different Archaea and Bacteria. Using comparative
tools, we identify close to a thousand highly-expressed candidate
noncoding RNAs. However, our analyses reveal that capacity to identify
noncoding RNA outputs is strongly dependent on phylogenetic
sampling. Surprisingly, and in stark contrast to protein-coding genes,
the phylogenetic window for effective use of comparative methods is
perversely narrow: aggregating public datasets only produced one
phylogenetic cluster where these tools could be used to robustly
separate unannotated noncoding RNAs from a null hypothesis of
transcriptional noise. Our results show that for the full potential of
transcriptomics data to be realized, a change in experimental design
is paramount: effective transcriptomics requires phylogeny-aware
sampling.

@@

Genome sequencing has transformed microbiology, offering unprecedented
insight into the physiology, biochemistry, and genetics of Bacteria
and Archaea \cite{Wu:2009,Rinke:2013,Loman:2012,Chun:2014}. Equally,
careful examination of transcriptional outputs has revealed that
bacterial and archaeal transcriptomes are remarkably complex
\cite{Sorek:2011}, containing a diverse array of noncoding RNA (ncRNA)
families, including sRNAs, sno-like sRNAs in Archaea, crRNAs and
riboswitches
\cite{Storz:2011,Dennis:2001,Horvath:2010,Breaker:2012,Cech:2014}. However,
our view of the microbial RNA world still derives from a narrow
sampling \cite{Pagani:2012} of microbial diversity. Additional bias
comes from the fact that many microbes are not readily culturable
\cite{Stewart:2012}. The development of metagenomics and initiatives
such as the Genomic Encyclopedia of Bacteria and Archaea (GEBA)
project have sought to redress these biases, generating genomes
spanning undersampled regions of the bacterial and archaeal phylogeny
\cite{Wu:2009}, and sequencing uncultured or unculturable species
through metagenomics \cite{Elkins:2008,Konneke:2005,Tyson:2004,Woyke:2006,Rinke:2013}.

A further source of bias in our genome-informed view of microorganisms
derives from a protein-centric approach to genome annotation. The
majority of genome sequences deposited in public databases carry
limited annotation of noncoding RNAs and cis-regulatory elements, yet
it is rapidly becoming clear that RNA is essential to our
understanding of molecular functioning in microbes \cite{mandin2013rna}.

The paucity of annotations is understandable, as RNA gene annotation
is non-trivial \cite{Freyhult:2007,Nawrocki:2009}. However, the
increasing roles for RNA uncovered through experimental and
bioinformatic study make illuminating this dark matter all the more
urgent. Among the remarkable discoveries made are: riboswitch-mediated
regulation \cite{Barrick:2007}, transcriptional termination by RNA
elements \cite{Gardner:2011,Chen:2013,Cambray:2013}, identification of
novel natural catalytic RNAs \cite{Roth:2014}, CRISPR-mediated
acquired immunity \cite{Barrangou:2007}, temperature-dependent gene
regulation \cite{Loh:2013}, and sno-like RNAs in Archaea
\cite{Omer:2000,Gardner:2010}. The Rfam database
\cite{Gardner:2011,Burge:2013} provides a valuable platform for
collating and characterising these and other families of noncoding
RNA. However, a recent comparative analysis \cite{Hoeppner:2012}
revealed that fewer than 7\% of RNA families within Bacteria and less
than 19\% in Archaea show a broad phylogenetic distribution (that is,
presence in at least 50\% of sequenced phyla). Crucially, that
analysis revealed that underlying genome sequencing biases were a
major contributor to this pattern, and that the wider genomic sampling
provided by the GEBA dataset did help improve identification of
broadly-conserved RNA families \cite{Hoeppner:2012}. Tools such as
RNA-seq \cite{Croucher:2010} and transposon insertion
sequencing \cite{van_Opijnen:2013,Barquist:2013,Barquist:2013a}
promise to complement comparative genomics tools for RNA family
discovery, and it may be possible to use a mix of data types in the
identification of RNA elements. However, to date, no systematic
analysis of available data has been undertaken, suggesting ncRNAs may
be hidden in the deluge of published data.

@@

To assess the relative contribution of noncoding RNAs and
protein-coding genes to transcriptional output, we collected all
publicly-available bacterial and archaeal RNA-seq datasets (available
as of {\bf August 2013}), spanning 37 species/strains and 400
datasets. For all datasets, we supplemented publicly available genome
annotations with screening for additional loci against Pfam and Rfam
databases, followed by manual identification of expressed unannotated
regions. This latter annotation yielded 922 expressed RNAs of
Unknown Function (RUFs).

@@

To assess whether highly expressed RUFs possess features commonly
associated with function, we employed three criteria: 1) evolutionary
conservation, 2) conservation of secondary structure, 3) evidence of
expression in more than one RNA-seq dataset. For this analysis, we
compared and ranked transcriptional outputs across species (see
Methods for details). Based on the relative rank across RNA-seq
datasets and the maximum phylogenetic distance observed across all
genomes, each transcript was classified as high, medium or low
expression, and high, medium or low conservation. This yielded a set
of highly expressed transcripts consisting of 162 Rfam families, 568
RUFs and 1429 Pfam families. As expected
\cite{Rocha:2003,Pal:2001,Drummond:2005}, conserved, highly expressed
outputs are dominated by protein-coding transcripts
(Figure~\ref{fig:2}B\&C). In contrast, transcripts that are highly
expressed but poorly conserved are primarily RUFs
(Figure~\ref{fig:2}A). Of the 568 RUFs identified, only 25 are
supported by all three criteria (conservation, secondary structure and
expression) (Figure~\ref{fig:2}D). A further 138 RUFs are supported by two criteria
(Figure~\ref{fig:2}D). Consequently, on these criteria, the vast majority of RUFs
appear indistinguishable from transcriptional noise. However, as these
RUFs are among the mostly highly expressed transcripts in public
RNA-seq data, we next considered whether our criteria were
sufficiently discriminatory to identify functional RNAs. It is well
established that not all functional RNAs exhibit conserved secondary
structure – antisense base pairing with a target is common, and does
not require intramolecular folding \cite{Gottesman:2011}. This
indicates that criterion 2 will apply to some, but not all functional
RNA elements. Criteria 1 and 3 both derive from comparative analysis:
criterion 1 requires an expressed RUF to be conserved in some other
genome, while criterion 2 requires an expressed RUF to be expressed in
another of the datasets in our study. We therefore sought to examine
how effective our comparative analyses are given that the available
data represent a small sample (transcriptomes from 37 species) and
given that biases in genome sampling across bacterial and archaeal
diversity impact comparative analysis of RNAs \cite{Hoeppner:2012}.

@@

Effective comparative analysis requires appropriate phylogenetic
distances between species under investigation \cite{Eddy:2005}. The
transcriptomes available for our analyses are limited in number and
phylogenetic diversity, and efforts to redress phylogenetic biases in
genome sequencing have focused on maximising phylogenetic diversity
across the tree of life \cite{Wu:2009,Rinke:2013}. Moreover,
underlying biases in genome sampling have been shown to affect the
assessment of ncRNA conservation \cite{Hoeppner:2012}. We therefore
considered how species selection impacts comparative analysis as a
tool for distinguishing signals of conservation in ncRNAs from
transcriptional noise.

To assess the effect of species selection on our capacity to identify
RNA families using comparative analysis, we first generated F84
phylogenetic distances between 2562 bacterial species and 154 archaeal
species using SSU rRNA sequences from each species (Methods). Next,
for each Rfam RNA family and Pfam protein family, we identified the
maximum phylogenetic distance between any two species that encode a
given family. We then calculated the fraction of conserved RNA and
protein families for a given phylogenetic distance.

This reveals a dramatic difference in evolutionary conservation of
Rfam and Pfam families (Figure~\ref{fig:3}). While 80\% of protein families are
still conserved at the broad evolutionary distances that separate
Bacteria and Archaea, the phylogenetic distance at which 80\% of RNA
families are conserved lies somewhere between the taxonomic levels of
genus and family (Figure~\ref{fig:3}). The explanation for this rapid decay of RNA
family conservation across long evolutionary time-scales is likely to
be a combination of the limited abilities of existing bioinformatic
tools to correctly align RNA sequences \cite{Gardner:2005} and genuine
rapid turnover of non-coding RNAs during evolution
\cite{Hoeppner:2012}.

These results in turn indicate that appropriate evolutionary distances
for optimising comparative analysis differ greatly for protein- and
RNA-coding genes. Figure~\ref{fig:3} confirms the utility of the GEBA sampling
strategy \cite{Wu:2009,Rinke:2013} for protein-coding gene
identification, since maximising phylogenetic diversity yields strong
identification of conserved protein-coding genes. In contrast, at the
largest phylogenetic distances, around 60\% of RNA families are
invisible to comparative analysis. These results define a ‘Goldilocks
Zone’ (an evolutionary distance neither too close nor too distant) for
RNA family comparative analysis.

To assess how phylogenetic sampling impacts comparative analysis of
transcriptome data, we mapped the pairwise distances between species
covered by the RNA-seq datasets in this study. Of the 506 possible
pairs (excluding Bacteria vs Archaea), only 11 are in the Goldilocks
Zone for RNA (phylogenetic distance between 0.0118 and 0.0542)
covering 9 species/strains. While five pairs of datasets are ‘too hot’
(i.e. too close phylogenetically), the remaining 490 comparisons are
‘too cold’ for effective comparative RNA analysis (Figure~\ref{fig:3}). The
datasets in the Goldilocks Zone span three distinct clades covering
five Enterobacteria, two Pseudomanada, and two Xanthomonada
(Figure~\ref{fig:3}). Subsequent analyses focused on the Enterobacterial clade, as
this was the only clade containing transcriptomes from multiple
species.

We next calculated the percentage of conserved RUFs for all
Enterobacterial species pairs. On average, 83\% of RUFs are conserved
across the Goldilocks Zone. The two E. coli strains are extremely
similar, and share 99\% of RUFs, suggesting that these species are too
similar for us to robustly separate expression of bona fide RNAs and
noise. While these outputs could be genuine RNAs, these strains are in
the ‘too hot’ region: if everything is conserved, comparative power is
lost. In contrast, only 12\% of RUFs are conserved between species
pairs in the ‘too cold’ region (spanning clades; Figure~\ref{fig:3}). Indeed, most
of the currently available datasets fall outside the Goldilocks Zone
for RNA discovery (Figure~\ref{fig:3}): of the 197 RUFs found in the datasets that
cover this region, only 19 show evidence of expression in another
transcriptome. This suggests that the low number of RUFs from Figure~\ref{fig:2}D
showing both conservation and expression in other AREBA species is
primarily a consequence of limited sampling. We also extracted a RUF
in the Enterobacterial clade that had homologs in the other four
species. In all cases, the homologs showed evidence of expression when
examining the RNA-seq data, and the five homologs show evidence of a
conserved secondary structure (Figure~\ref{fig:3}).

@@

This use of transcriptome data clearly improves our capacity to
identify noncoding outputs: applying three criteria (sequence
conservation, conservation of secondary structure, and expression in
multiple species) we have identified 163 high-confidence expressed
RUFs from public data (Figure~\ref{fig:2}). An additional 405 RUFs are highly
expressed across the transcriptomes we have examined, yet these do not
show clear signs of sequence or structural conservation in other
sequenced genomes. Given their high expression level, these seem
unlikely to be transcriptional noise. Some may represent technical
artefacts, but many could be bona fide lineage-specific ncRNAs with
potentially novel functions.

@@

\begin{CJK}{UTF8}{min}
Given that isolation, cultivation and study of individual bacterial
and archaeal strains can be extremely challenging (Stewart 2012),
successful phylogeny-informed comparative RNA-seq will be a demanding
endeavour, requiring complex sets of expertise spanning advanced
culturing and isolation techniques, functional genomics capability and
RNA bioinformatics. This places such a project beyond the reach of
most individual labs. We therefore propose that comprehensive
resolution of the comparative RNA-seq problem can best be resolved via
a community-driven initiative: in recognition of the success of the
GEBA project, we have dubbed this An RNA Encyclopedia of Bacteria and
Archaea (AREBA). The appropriateness of this acronym will be
especially clear to Japanophones, as, in Japanese, the phrase ‘areba’
(あれば) translates to ‘if there’.
\end{CJK}

@@

All available bacterial and archaeal genomes were downloaded from the
European Nucleotide Archive (ENA) (2,562 and 154 genomes,
respectively) \cite{Cochrane:2013}. RNA-seq datasets published as of
March 2013 were collected, spanning 37 species/strains, 44 experiments
and 413 lanes of sequencing data (Table S1). Most of these datasets
are generated on the Illumina platform \cite{Shendure:2005}, with a
few lanes from the SOLiD platform \cite{Valouev:2008} and 454
platform \cite{Margulies:2005}. Where possible, FastQ files were
downloaded, scanned for residual adapter sequences using
AdapterRemoval \cite{Lindgreen:2012}, and mapped to the reference genome
using Bowtie2 \cite{Langmead:2012} for Illumina and 454 data
and BFAST \cite{Homer:2009} for SOLiD data.

@@

Homologs of the identified RUFs were found in all the downloaded
genomes using nhmmer \cite{Wheeler:2013} in an iterative fashion:
First, the RUF sequence alone is used in the scan; then, all hits with
$E-value<0.001$ are included and a HMM is built, and this is iterated
5 times. The alignments from the RUF homology search were analyzed
further by investigating the potential for secondary structure
formation using RNAz \cite{Gruber:2010} and alifoldz
\cite{Washietl:2004}; the protein coding potential using RNAcode
\cite{Washietl:2011}, and overlaps between potential RUF homologs in
other species and all the annotations in the respective genomes as
described above.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{figures/figure2.pdf}
\end{center}
\caption{ {\bf Many ncRNAs and RUFs are highly expressed
    but show limited conservation across represented species.} {\bf
    A-C:} Both known Rfam families and the RUFs identified in this
  analysis are often highly expressed
  transcripts. In contrast to protein-coding transcripts (blue), where
  highly-expressed transcripts are well-conserved, the opposite is
  true of many non-coding RNA elements (Rfam, red; RUFs,
  black). Notably, the greatest proportion of highly expressed
  Rfam-annotated RNA elements show a narrow evolutionary
  distribution. This is also reflected in the RUFs identified in this
  study. {\bf D:} Venn diagram of the 568 highly expressed RUFs. Each
  RUF was analysed to look for evidence of secondary structure
  formation, level of conservation, and evidence of expression in at
  least one other RNA-seq dataset. All RUFs showing expression in
  other AREBA species must of course be conserved in at least two
  species, so the figure also shows that 219 highly expressed RUFs are
  conserved across a limited phylogenetic distance only.}
\label{fig:2}
\end{figure}
%pdftops -eps figures/figure2.pdf figures/figure2.eps

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{figures/figure4.pdf}
\end{center}
\caption{ {\bf Comparative analysis of RNA-seq datasets in the
    'Goldilocks Zone' is a powerful approach for identifying 'RNAs of
    Unknown Function'.} {\bf (A)} Ten species with available RNA-seq data
  and phylogenetic distances in the Goldilocks Zone (GZ) (Figure 3)
  have been identified. The maximum likelihood tree from a SSU rRNA
  alignment shows the relationships between these taxa. They fall into
  three clades, containing members of the families: Enterobacteriaceae
  and Xanthomonadaceae, and the genus: Pseudomonas. Nodes connecting
  taxa within the GZ are coloured gold, taxa that are too close are
  coloured red and those that are too divergent are coloured cyan. {\bf (B)}
  The predicted secondary structure for an exemplar RNA of Unknown
  Function (RUF) chosen from the Enterobacteriaceae data. This RUF is
  highly expressed in each species (C), shows co-variation within the
  second hairpin (highlighted in green, structure neutral variation is
  highlighted in blue, highly conserved regions are highlighted in
  pink), it conserves a tetraloop of the GNRA or UNCG types and there
  have been two independent insertions of hairpins in the S. enterica
  and K. pneumoniae within the first hairpin. {\bf (C)} The expression
  levels inferred from RNA-seq in the locus containing the RUF
  illustrated in B. The locus contains a ncRNA (red: SraB) and three
  protein coding genes (blue: yceD, rpmF and plsX) and the RUF
  (red). For each nucleotide in the locus the total number of reads
  that map there was computed, these are illustrated in the heatmap;
  Darker colours indicate high relative expression, lighter colours
  indicate low expression and white indicates a gap in the alignment
  of the sequences for this locus from the five Enterobacteriaceae.}
\label{fig:4}
\end{figure}
%pdftops -eps figures/figure4.pdf figures/figure4.eps

>>>>

Noncoding RNAs are increasingly recognized as integral to a wide range
of biological processes, including translation, gene regulation,
host-pathogen interactions and environmental sensing. While genomics
is now a mature field, our capacity to identify noncoding RNA elements
in bacterial and archaeal genomes is hampered by the difficulty of \emph{de
novo} identification. The emergence of new technologies for
characterizing transcriptome outputs, notably RNA-seq, are improving
noncoding RNA identification and expression quantification. However, a
major challenge is to robustly distinguish functional outputs from
transcriptional noise. To establish whether annotation of existing
transcriptome data has effectively captured all functional outputs, we
analysed over 400 publicly available RNA-seq datasets spanning 37 different
Archaea and Bacteria. Using comparative tools, we identify close to a
thousand highly-expressed candidate noncoding RNAs. However, our
analyses reveal that capacity to identify noncoding RNA outputs is
strongly dependent on phylogenetic sampling. Surprisingly, and in
stark contrast to protein-coding genes, the phylogenetic window for
effective use of comparative methods is perversely narrow: aggregating
public datasets only produced one phylogenetic cluster where these
tools could be used to robustly separate unannotated noncoding RNAs
from a null hypothesis of transcriptional noise. Our results show that
for the full potential of transcriptomics data to be realized, a
change in experimental design is paramount: effective transcriptomics
requires phylogeny-aware sampling.

@@

Genome sequencing has transformed microbiology, offering unprecedented
insight into the physiology, biochemistry, and genetics of Bacteria
and Archaea \cite{Wu:2009,Rinke:2013,Loman:2012,Chun:2014}. Equally,
careful examination of transcriptional outputs has revealed that
bacterial and archaeal transcriptomes are remarkably complex
\cite{Sorek:2011}, and include regulation, post-transcriptional
modification and genome defense processes
\cite{Storz:2011,Dennis:2001,Horvath:2010,Breaker:2012,Cech:2014}. However,
our view of the microbial RNA world still derives from a narrow
sampling of microbial diversity \cite{Pagani:2012}. Additional bias
comes from the fact that many microbes are not readily culturable
\cite{Stewart:2012}. The development of metagenomics and initiatives
such as the Genomic Encyclopedia of Bacteria and Archaea (GEBA)
project have sought to redress these biases, generating genomes
spanning undersampled regions of the bacterial and archaeal phylogeny
\cite{Wu:2009}, and sequencing uncultured or unculturable species
through metagenomics
\cite{Elkins:2008,Konneke:2005,Tyson:2004,Woyke:2006,Rinke:2013}.

A further source of bias in our genome-informed view of microbes
derives from a protein-centric approach to genome annotation. The
majority of genome sequences deposited in public databases carry
limited annotation of noncoding RNAs and cis-regulatory elements, yet
it is rapidly becoming clear that RNA is essential to our
understanding of molecular functioning in microbes \cite{mandin2013rna}.

The paucity of annotations is understandable, as RNA gene annotation
is non-trivial \cite{Freyhult:2007,Nawrocki:2009}. However, the
increasing roles for RNA uncovered through experimental and
bioinformatic study make illuminating this ``dark matter'' all the more
urgent. Among the remarkable discoveries made are: riboswitch-mediated
regulation \cite{Barrick:2007,Breaker:2012}, transcriptional
termination by RNA elements
\cite{von_Hippel:1998,Gardner:2011,Santangelo:2011}, identification of
novel natural catalytic RNAs
\cite{Kruger:1982,Guerrier-Takada:1983,Winkler:2004,Roth:2014},
CRISPR-mediated acquired immunity \cite{Barrangou:2007,Brouns:2008},
temperature-dependent gene regulation \cite{Narberhaus:2006,Loh:2013},
and sno-like RNAs in Archaea
\cite{Omer:2000,Gaspin:2000,Gardner:2010}. The Rfam database
\cite{Gardner:2011,Burge:2013} provides a valuable platform for
collating and characterising these and other families of noncoding
RNA. However, a recent comparative analysis \cite{Hoeppner:2012}
revealed that fewer than 7\% of RNA families within Bacteria and less
than 19\% in Archaea show a broad phylogenetic distribution (that is,
presence in at least 50\% of sequenced phyla). Crucially, that
analysis revealed that underlying genome sequencing biases were a
major contributor to this pattern, and that the wider genomic sampling
provided by the GEBA dataset \cite{Wu:2009} did help improve
identification of broadly-conserved RNA families
\cite{Hoeppner:2012}. Tools such as RNA-seq \cite{Croucher:2010} and
transposon insertion sequencing
\cite{van_Opijnen:2013,Barquist:2013,Barquist:2013a} promise to
complement comparative genomics tools for RNA family discovery, and it
may be possible to use a mix of data types in the identification of
RNA elements. However, to date, no systematic analysis of available
data has been undertaken, suggesting ncRNAs may be hidden in the
deluge of published data.

@@

To assess the relative contribution of noncoding RNAs and
protein-coding genes to transcriptional output, we collected all
publicly-available bacterial and archaeal RNA-seq datasets (available
as of August 2013), spanning 37 species/strains and 413 datasets. For
all datasets, we supplemented publicly available genome annotations
with screening for additional loci against the Pfam and Rfam databases
\cite{Punta:2012,Finn:2014,Gardner:2011,Burge:2013}, followed by
manual identification of expressed unannotated regions. This latter
annotation yielded 922 expressed RNAs of Unknown Function (RUFs)
\cite{McCutcheon:2003}.

@@

To assess whether highly expressed RUFs possess features commonly
associated with function, we employed three criteria: 1) evolutionary
conservation, 2) conservation of secondary structure, 3) evidence of
expression in more than one RNA-seq dataset. For this analysis, we
compared and ranked transcriptional outputs across species (see
Methods for details). Based on the relative rank across RNA-seq
datasets and the maximum phylogenetic distance observed across all
genomes, each transcript was classified as high, medium or low
expression, and high, medium or low conservation. This yielded a set
of highly expressed transcripts consisting of 162 Rfam families, 568
RUFs and 1429 Pfam families. As expected
\cite{Rocha:2003,Pal:2001,Drummond:2005}, conserved, highly expressed
outputs are dominated by protein-coding transcripts
(Figure~\ref{fig:2}B\&C). In contrast, transcripts that are highly
expressed but poorly conserved are primarily RUFs
(Figure~\ref{fig:2}A). Of the 568 RUFs identified, only 25 are
supported by all three conservative criteria (conservation, secondary
structure and expression) (Figure~\ref{fig:2}D), a further 138 RUFs
are supported by two criteria (Figure~\ref{fig:2}D). Consequently, on
these criteria, the vast majority of RUFs appear indistinguishable
from transcriptional noise. However, as these RUFs are among the
mostly highly expressed transcripts in public RNA-seq data, we next
considered whether our criteria were sufficiently discriminatory to
identify functional RNAs. It is well established that not all
functional RNAs exhibit conserved secondary structure – antisense base
pairing with a target is common, and does not require intramolecular
folding \cite{Gottesman:2011}. This indicates that criterion 2 will
apply to some, but not all functional RNA elements. Criteria 1 and 3
both derive from comparative analysis: criterion 1 requires an
expressed RUF to be conserved in some other genome, while criterion 2
requires an expressed RUF to be expressed in another of the datasets
in our study. We therefore sought to examine how effective our
comparative analyses are given that the available data represent a
small sample (transcriptomes from 37 species) and given that biases in
genome sampling across bacterial and archaeal diversity impact
comparative analysis of RNAs \cite{Hoeppner:2012}.

@@

Effective comparative analysis requires appropriate phylogenetic
distances between species under investigation \cite{Eddy:2005}.  For
discovery of protein-coding gene families, maximising phylogenetic
diversity across the tree of life has proven very effective
\cite{Wu:2009,Rinke:2013}. For non-coding RNA, underlying biases in
genome sampling do affect the assessment of ncRNA conservation, and
adding phylogenetic diversity improves the identification of broadly
conserved ncRNA families \cite{Hoeppner:2012}. However, few ncRNAs
appear conserved across broad evolutionary distances
\cite{Hoeppner:2012}.
We have therefore considered how species selection impacts comparative
analysis as a tool for the identification of conserved ncRNAs.

To assess the effect of species selection on our capacity to identify
RNA families using comparative analysis, we first generated F84
phylogenetic distances between 2562 bacterial species and 154 archaeal
species using SSU rRNA sequences from each species (see Methods for
details). Next, for each Rfam RNA family and Pfam protein family, we
identified the maximum phylogenetic distance between any two species
that encode a given family. We then calculated the fraction of
conserved RNA and protein families for a given phylogenetic distance.

This reveals a dramatic difference in evolutionary conservation of
Rfam and Pfam families (Figure~\ref{fig:3}). While 80\% of protein families are
still conserved at the broad evolutionary distances that separate
Bacteria and Archaea, the phylogenetic distance at which 80\% of RNA
families are conserved lies somewhere between the taxonomic levels of
genus and family (Figure~\ref{fig:3}). The explanation for this rapid decay of RNA
family conservation across long evolutionary time-scales is likely to
be a combination of the limited abilities of existing bioinformatic
tools to correctly align RNA sequences \cite{Gardner:2005} and 
rapid turnover of non-coding RNAs during evolution
\cite{Hoeppner:2012}.

These results in turn indicate that appropriate evolutionary distances
for optimal comparative analysis differ greatly for protein- and
RNA-coding genes. Figure~\ref{fig:3} confirms the utility of the GEBA sampling
strategy \cite{Wu:2009,Rinke:2013} for protein-coding gene
identification, since maximising phylogenetic diversity permits effective
identification of conserved protein-coding genes. In contrast, at the
largest phylogenetic distances, around 60\% of RNA families are
invisible to comparative analysis. These results define a ‘Goldilocks
Zone’ (an evolutionary distance neither too close nor too distant) for
ncRNA analysis through comparative analysis.

In order to assess the potential for existing RNA-seq data to be used
for ncRNA analysis, we mapped the pairwise distances between species
covered by the RNA-seq datasets in this study. Of the 506 possible
pairs (excluding Bacteria vs Archaea), only 11 are in the Goldilocks
Zone for RNA (phylogenetic distance between 0.0118 and 0.0542)
covering 9 species/strains. While five pairs of datasets are ‘too hot’
(i.e. too close phylogenetically), the remaining 490 comparisons are
‘too cold’ for effective comparative RNA analysis
(Figure~\ref{fig:4}). The datasets in the Goldilocks Zone span three
distinct clades covering five Enterobacteria, two Pseudomanada, and
two Xanthomonada (Figure~\ref{fig:4}). Subsequent analyses focused on
the Enterobacterial clade, as this was the only clade containing
transcriptomes from multiple species.

We next calculated the percentage of conserved RUFs for all
Enterobacterial species pairs. On average, 83\% of RUFs are conserved
across the Goldilocks Zone. The two \emph{E. coli} strains are extremely
similar, and share 99\% of their RUFs, suggesting that these species
are too similar for us to robustly separate expression of \emph{bona fide}
RNAs and noise. While these outputs could be genuine RNAs, these
strains are in the ‘too hot’ region, meaning if everything is
conserved, comparative power is lost. In contrast, only 12\% of RUFs
are conserved between species pairs in the ‘too cold’ region (spanning
clades; Figure~\ref{fig:3}) and of the 197 RUFs found through
comparative analysis of transcriptomes within the Goldilocks Zone,
only 19 show evidence of expression in another transcriptome outside
of this zone. This suggests that the low number of RUFs from
Figure~\ref{fig:2}D showing both conservation and expression is
primarily a consequence of limited sampling.  We also extracted a RUF
in the Enterobacterial clade that had homologs in the other four
species.  That said, mining RNA-seq data within the Goldilocks Zone
permits a higher confidence in the identification of novel ncRNAs. One
such example of this is illustrated in (Figure~\ref{fig:4}B\&C). This
RUF exhibits sequence and secondary structure conservation and is
expressed at high levels across all five Enterobacterial
transcriptomes in our dataset.

@@

This use of transcriptome data clearly improves our capacity to
identify noncoding outputs: applying three criteria (sequence
conservation, conservation of secondary structure, and expression in
multiple species) we have identified 163 high-confidence expressed
RUFs from public data (Figure~\ref{fig:2}). An additional 405 RUFs are highly
expressed across the transcriptomes we have examined, yet these do not
show clear signs of sequence or structural conservation in other
sequenced genomes. Given their high expression level, these seem
unlikely to be transcriptional noise. Some may represent technical
artefacts, but many could be \emph{bona fide} lineage-specific ncRNAs with
potentially novel functions.

@@

\begin{CJK}{UTF8}{min}
Given that isolation, cultivation and study of individual bacterial
and archaeal strains can be extremely challenging \cite{Stewart:2012}
successful phylogeny-informed comparative RNA-seq will be a demanding
endeavour, requiring complex sets of expertise spanning advanced
culturing and isolation techniques, functional genomics capability and
RNA bioinformatics. This places such a project beyond the reach of
most individual labs. We therefore propose that comprehensive
resolution of the comparative RNA-seq problem can best be resolved via
a community-driven initiative: in recognition of the success of the
GEBA project, we have dubbed this An RNA Encyclopedia of Bacteria and
Archaea (AREBA). The appropriateness of this acronym will be
especially clear to Japanophones, as, in Japanese, the phrase ‘areba’
(あれば) translates to ‘if there’.
\end{CJK}

@@

All available bacterial and archaeal genomes were downloaded from the
European Nucleotide Archive (ENA) (2,562 and 154 genomes,
respectively) \cite{Cochrane:2013}. RNA-seq datasets published as of
March 2013 were collected, spanning 37 species/strains, 44 experiments
and 413 lanes of sequencing data (Table S1). Most of these datasets
are generated on the Illumina platform \cite{Shendure:2005}, with a
few lanes from the SOLiD platform \cite{Cloonan:2008} and the 454
platform \cite{Margulies:2005}. Where possible, FastQ files were
downloaded, scanned for residual adapter sequences using
AdapterRemoval \cite{Lindgreen:2012}, and mapped to the reference genome
using Bowtie2 \cite{Langmead:2012} for Illumina and 454 data
and BFAST \cite{Homer:2009} for SOLiD data.

@@

Homologs of the identified RUFs were found in all the downloaded
genomes using nhmmer \cite{Wheeler:2013} in an iterative fashion:
First, the RUF sequence alone was used in the scan; then, all hits
with $E-value<0.001$ were included and a HMM built. This was iterated
5 times. The alignments from the RUF homology search were analyzed
further by investigating the potential for secondary structure
formation using RNAz \cite{Gruber:2010} and alifoldz
\cite{Washietl:2004}. Protein coding potential of the RUFs was
assessed using RNAcode \cite{Washietl:2011}. Overlaps between
potential RUF homologs in other species and all the annotations in the
respective genomes was also assessed.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{figures/figure2.pdf}
\end{center}
\caption{ {\bf Many ncRNAs and RUFs are highly expressed
    but show limited conservation across represented species.} {\bf
    A-C:} Both known Rfam families and the RUFs identified in this
  analysis are often highly expressed
  transcripts. In contrast to protein-coding transcripts (blue), where
  highly-expressed transcripts are well-conserved, the opposite is
  true of many non-coding RNA elements (Rfam, red; RUFs,
  black). Notably, the greatest proportion of highly expressed
  Rfam-annotated RNA elements show a narrow evolutionary
  distribution. This is also reflected in the RUFs identified in this
  study. {\bf D:} Venn diagram of the 568 highly expressed RUFs. Each
  RUF was analysed to look for evidence of secondary structure
  formation, level of conservation, and evidence of expression in at
  least one other RNA-seq dataset. All RUFs showing expression in
  other species must of course be conserved in at least two
  species, so the figure also shows that 219 highly expressed RUFs are
  conserved across a limited phylogenetic distance only.}
\label{fig:2}
\end{figure}
%pdftops -eps figures/figure2.pdf figures/figure2.eps

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{figures/figure4.pdf}
\end{center}
\caption{ {\bf Comparative analysis of RNA-seq datasets in the
    'Goldilocks Zone' is a powerful approach for identifying 'RNAs of
    Unknown Function'.} {\bf (A)} Ten species with available RNA-seq data
  and phylogenetic distances in the Goldilocks Zone (GZ) (Figure 3)
  have been identified. The maximum likelihood tree from a SSU rRNA
  alignment shows the relationships between these taxa. They fall into
  three clades, containing members of the families: Enterobacteriaceae
  and Xanthomonadaceae, and the genus: Pseudomonas. Nodes connecting
  taxa within the GZ are coloured gold, taxa that are too close are
  coloured red and those that are too divergent are coloured cyan. {\bf (B)}
  The predicted secondary structure for an exemplar RNA of Unknown
  Function (RUF) chosen from the Enterobacteriaceae data. This RUF is
  highly expressed in each species (C), shows co-variation within the
  second hairpin (highlighted in green, structure neutral variation is
  highlighted in blue, highly conserved regions are highlighted in
  pink), it conserves a tetraloop of the GNRA or UNCG types and there
  have been two independent insertions of hairpins in \emph{S. enterica}
  and \emph{K. pneumoniae} within the first hairpin. {\bf (C)} The expression
  levels inferred from RNA-seq in the locus containing the RUF
  illustrated in B. The locus contains a ncRNA (red: SraB) and three
  protein coding genes (blue: yceD, rpmF and plsX) and the RUF
  (red). For each nucleotide in the locus the total number of reads
  that map there was computed, these are illustrated in the heatmap;
  Darker colours indicate high relative expression, lighter colours
  indicate low expression and white indicates a gap in the alignment
  of the sequences for this locus from the five Enterobacteriaceae.}
\label{fig:4}
\end{figure}
%pdftops -eps figures/figure4.pdf figures/figure4.eps
